Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-09-17 paper/main.tex
finished projects section

Thus, rather than constructing fully separate Python environments from scratch, 
%Davos projects serve to fill in the gap between the user's existing environment and the dependencies of their corresponding notebooks. 
Davos projects work by supplementing the user's existing environment with any additional packages (or specific package versions) needed to satisfy the dependencies of their corresponding notebooks.
In some cases, this might include every package smuggled into a notebook (e.g., if the notebook is run inside a freshly created, empty virtual environment). 
In other cases, the user's environment may already provide all required packages, and the notebook's project directory will be unused (in which case it will be deleted automatically when the notebook kernel is shut down). 
But regardless of the extent to which the existing environment is augmented, Davos's project system ensures that all smuggled packages are installed locally and loaded successfully at runtime, while the contents of the user's Python environment are never altered. 
Additionally, because \texttt{smuggle} statements in a given notebook are evaluated every time it is run, this system also ensures that the notebook's requirements will continue to be satisfied even if the user's Python environment changes. 
For example, suppose a user has \texttt{NumPy}~\cite{HarrEtal20} v1.24.3 installed in their current Python environment and runs a Davos-enhanced notebook that smuggles \texttt{NumPy} with ``\texttt{numpy==1.24.3}'' specified in an onion comment (see Sec.~\ref{subsec:onion}). 
Since the user's existing version of the package satisfies this requirement, Davos will happily load it into the notebook. 
But if the user later upgrades their environment's \texttt{NumPy} version to v1.25.0 (perhaps as a result of installing a different package that depends on it) and subsequently reruns this notebook, the local version will longer satisfy this requirement, so Davos will install \texttt{NumPy} v1.24.3 into the notebook's project directory and load that version instead. 
From then on, any further changes to the user's \texttt{Numpy} installation would have no effect on Davos's behavior in this particular notebook, as a satisfactory version now exists in its project directory. (If the version specified in the onion comment were changed, Davos would install it in the project directory in place of the previous version.) 
For efficiency, Davos projects will generally avoid duplicating dependencies already satisfied by the user's Python environment. 
However, if desired, adding \texttt{pip}'s \texttt{--ignore-installed} flag to an onion comment in the notebook will cause the smuggled package to be installed into the project directory whether or not it already exists locally.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADD THIS EITHER TO START OF PROJECTS SUBSUBSECTION OR IN IMPACT SECTION   %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%A common way of avoiding this is to create a virtual environment in which to run the notebook, and instruct anyone with whom the notebook is shared to do the same. While effective, this added requirement introduces additional 
%
%
%(or group of related notebooks) that require different 
%
%
%\begin{itemize}
%  \item{common solution is to use a virtual environment}
%  \item{this introduces complexity}
%  \item{davos's solution is projects}
%\end{itemize}

@@

%Because Davos can install new packages, running the code in a
%Davos-enhanced notebook might (in principle) affect the behavior of
%\textit{other} Python-based software (e.g., other notebooks, scripts, etc.) by
%altering which packages are installed in the runtime environment. This could
%lead to undesired consequences. For example, suppose Person A develops a
%notebook (Notebook A) for their research project. We will assume that Notebook
%A does not use Davos to manage project dependencies.  If Person A
%runs a Davos-enhanced Notebook B, e.g., sent by another developer,
%might this unexpectedly affect the behavior of Notebook A?

%To prevent unwanted
%changes to the user's Python environment, Davos incorporates its
%own virtual environment-like scheme for isolating packages it installs. When Davos
%is imported, a new virtual environment (folder) is created automatically. The
%folder's name may be customized to support multi-notebook projects. Any
%\texttt{smuggle}d packages that were not available in the notebook's runtime
%environment are installed to the current project folder. The runtime
%environment remains unaffected by Davos's behavior (see
%Sec.~\ref{subsec:projects}).
%By default, projects are notebook-specific, but can also be shared by multiple notebooks, and can be managed interactively from within a Davos-enhanced notebook.

%%%%%%%% JEREMY VERSION %%%%%%%%%
%Installing new packages in a notebook using standard approaches (e.g., system commands) affect the
%runtime environment. This could lead to undesired behaviors. For example, running a notebook that
%installs new packages in the user’s primary system environment might alter their main system installation
%and/or containerized environment in unexpected ways (e.g., changing package versions, causing conflicts
%with other packages, etc.). To protect against undesired changes to the runtime environment, Davos
%incorporates its own virtual environment for managing packages it installs. When Davos is imported,
%a new virtual environment (folder) is created automatically. The folder’s name may be customized to
%support multi-notebook projects. Any smuggled packages that were not available in the notebook’s
%runtime environment are installed to the current project folder. The runtime environment remains unaffected
%by Davos’s behavior (see Sec. 2.2.3).

We implemented a ``project'' system in Davos to protect against the
above scenario. By default, importing Davos creates a new project
folder in the user's home directory (contained within a hidden \texttt{.davos}
folder). The default project name is computed to uniquely identify each
notebook according to its filename and path. Any packages that were not
originally available in the notebook's runtime environment are installed to the
notebook's project directory. When external libraries are \texttt{smuggle}d,
Davos temporally appends the current project directory to the search
path. Because the user's system path remains unchanged, and because none of the
runtime environment's packages are altered, the user's system and runtime
environment remain unaffected (aside from installing the Davos package
itself to the runtime environment).

Each notebook's project may be customized by setting \texttt{davos.project} to
any string that can be used as a valid folder name in the user's operating
system. By customizing the project name, users can build multi-notebook
projects that share the same core set of dependencies without needing to
duplicate each package for each notebook in the project.

Finally, if the user \textit{does} wish to modify their runtime environment,
this may be done by setting \texttt{davos.project} to \texttt{None}. Doing so
will cause any packages installed by Davos to affect the user's
runtime environment.  This is generally not recommended, as it can lead to
unintended consequences for other code that shares the runtime environment.

@@

@@

>>>>

Thus, rather than constructing fully separate Python environments from scratch, 
%Davos projects serve to fill in the gap between the user's existing environment and the dependencies of their corresponding notebooks. 
Davos projects work by supplementing the user's existing environment with any additional packages (or specific package versions) needed to satisfy the dependencies of their corresponding notebooks.
In some cases, this might include every package smuggled into a notebook (e.g., if the notebook is run inside a freshly created, empty virtual environment). 
In other cases, the user's environment may already provide all required packages, and the notebook's project directory will be unused (in which case it will be deleted automatically when the notebook kernel is shut down). 
But regardless of the extent to which the existing environment is augmented, Davos's project system ensures that all smuggled packages are installed locally and loaded successfully at runtime, while the contents of the user's Python environment are never altered. 
Additionally, because \texttt{smuggle} statements in a given notebook are evaluated every time it is run, this system also ensures that the notebook's requirements will continue to be satisfied even if the user's Python environment changes. 
For example, suppose a user has \texttt{NumPy}~\cite{HarrEtal20} v1.24.3 installed in their current Python environment and runs a Davos-enhanced notebook that smuggles \texttt{NumPy} with ``\texttt{numpy==1.24.3}'' specified in an onion comment (see Sec.~\ref{subsec:onion}). 
Since the user's existing version of the package satisfies this requirement, Davos will happily load it into the notebook. 
But if the user later upgrades their environment's \texttt{NumPy} version to v1.25.0 (perhaps as a result of installing a different package that depends on it) and subsequently reruns this notebook, the local version will longer satisfy this requirement, so Davos will install \texttt{NumPy} v1.24.3 into the notebook's project directory and load that version instead. 
From then on, any further changes to the user's \texttt{Numpy} installation would have no effect on Davos's behavior in this particular notebook, as a satisfactory version now exists in its project directory. 
(If the version specified in the onion comment were changed, Davos would install it in the project directory in place of the previous version.) 
For efficiency, Davos projects will generally avoid duplicating dependencies already satisfied by the user's Python environment. 
However, if desired, adding \texttt{pip}'s \texttt{--ignore-installed} flag to an onion comment in the notebook will cause Davos to install the smuggled package into the project directory whether or not it already exists locally.

By default, each Davos-enhanced notebook will create and use its own notebook-specific project named for the absolute path to the notebook file. 
However, before smuggling its required packages, a notebook may be set to instead use an arbitrarily named, notebook-agnostic project by assigning any string (that is a valid Unix directory name) to \texttt{davos.project}.
This provides a convenient way for multiple related notebooks that share a common set of requirements to use the same Davos project, by setting \texttt{davos.project} to the same string in each one. 
%This provides a convenient way to share a single Davos project across multiple related notebooks that share a common set of requirements, by setting \texttt{davos.project} to the same string in each one. 
It is also possible (though not recommended) to disable Davos's project system entirely and install smuggled packages directly into the user's Python environment by setting \texttt{davos.project} to \texttt{None}.

When accessed (unless its value has been set to \texttt{None}), \texttt{davos.project} will return a \texttt{Project} object that represents the project used by the current notebook (strings assigned to \texttt{davos.project} are converted to \texttt{Project}s internally). 
This object supports methods for interacting with the current project, including locating its directory on the file system, listing all installed packages' names and versions, changing the project's name, and deleting its contents altogether.
\texttt{Project}s can also be created and managed programmatically, and Davos provides additional top-level functions for viewing and working with all existing projects.

@@

@@

% \subsubsection{Other top-level Davos functions}\label{subsec:toplevel}

@@

\textcolor{red}{\textbf{TODO: add note about default/fallback project for non-traditional notebook interfaces}}
